Short-time expansion of Heisenberg operators in open collective quantum spin systems

### 2019-04-28 soc_squeezing/collective_paper/collective_spin.tex
changes to figures and associated commentary in collective spin paper

\title{Short-time simulations of open collective quantum spin systems}

@@

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $\hat S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is
the directionally-unbiased Ramsey squeezing parameter determined by
the maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$
over that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\lvert\braket{\uv S}\rvert^2}
  \min_{\uv n\perp\bk{\uv S}} \Braket{\p{\uv S\cdot\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\uv S\equiv\p{\hat S_\x,\hat S_\y,\hat S_\z}$ is a collective
spin operator-valued vector, and the minimization is performed over
all unit vectors $\uv n$ in the plane orthogonal to the mean spin
vector $\bk{\uv S}$.  This squeezing parameter is entirely determined
by one- and two-spin correlators of the form $\bk{\hat S_\a}$ and
$\bk{\hat S_\a \hat S_\b}$.  For the unitary dynamics discussed in
this work, these correlators are obtainable via exact simulations of
quantum dynamics in the $\p{N+1}$-dimensional Dicke manifold of states
$\set{\ket{S,m}}$ with net spin $S$ and spin projection $m$ onto the
$z$ axis, i.e.~with $\bk{S,m|\uv S^2|S,m}=S\p{S+1}$ and
$\bk{S,m|\hat S_\z|S,m}=m$ for $m\in\set{-S,-S+1,\cdots,S}$.  In the
presence of single-spin or collective decoherence, meanwhile, these
correlators are obtainable with the collective-spin quantum trajectory
Monte Carlo method developed in ref.~\cite{zhang2018montecarlo}.  In
this work, these exact and quantum trajectory simulations will be used
to benchmark the TST expansion in Eq.~\eqref{eq:TST}.

\begin{figure}
  \centering
  \subfloat[Squeezing with unitary dynamics
  \label{fig:squeezing_coherent}]
  {\includegraphics{coherent.pdf}} \\
  \subfloat[Squeezing with decoherence:
  $\gamma_-=\gamma_+=\gamma_\z=\chi$
  \label{fig:squeezing_incoherent}]
  {\includegraphics{decoherence_weak.pdf}}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under \sref{fig:squeezing_coherent} unitary and
    \sref{fig:squeezing_incoherent} non-unitary dynamics, computed
    using exact methods (solid lines), quantum trajectory simulations
    (dots), and the TST expansion in Eq.~\eqref{eq:TST} with $M=35$
    (darker dashed lines).  Solid circles mark the times at which the
    TST expansion gives an unphysical result with $\xi^2<0$.}
  \label{fig:benchmarking}
\end{figure}

Figure \ref{fig:benchmarking} compares the squeezing parameter $\xi^2$
for $N=10^4$ spins initially in the state $\ket{\X}$ evolved under the
Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}, as computed via
both benchmarking simulations and the TST expansion in
Eq.~\eqref{eq:TST} with $M=35$.  Squeezing is shown for both unitary
dynamics (Figure \ref{fig:squeezing_coherent}), as well as non-unitary
dynamics in the presence of spontaneous decay, excitation, and
dephasing of individual spins at rates $\chi$ (Figure
\ref{fig:squeezing_incoherent}), respectively described by the sets of
jump operators $\A_\a\equiv\set{\hat s_\a^{(j)}}$ with corresponding
decoherence rates $\gamma_\a=\chi$ for $\a\in\set{-,+,\z}$.  The
results shown in Figure \ref{fig:benchmarking} were computed in a
rotated basis with
$\p{\hat s_\z,\hat s_\x}\to\p{\hat s_\x,-\hat s_\z}$ and
$\ket\X\to\ket{-\Z}\equiv\ket\dn^{\otimes N}$ (together with
appropriate transformations of the Hamiltonian and jump operators);
the only effect of this rotation on the results presented in Figure
\ref{fig:benchmarking} is to prolong the time for which the TST
expansion agrees with the benchmarking simulations.  The reason for
different results in rotated basis has to do with the breakdown of the
TST expansion, which we discuss further below.

The main lesson from Figure \ref{fig:benchmarking} is that the TST
expansion gives essentially exact results right up until a sudden and
drastic departure that can be diagnosed by inspection.  This drastic
departure occurs because the TST expansion neglects high-weight
(i.e.~large-$\abs{\v m}$) operators $\hat\S_{\v m}$ whose
contributions to a Heisenberg operator $\hat\O\p{t}$ of interest
eventually become non-negligible.  For sufficiently large times $t$,
these individual contributions may generally be large compared to the
value of $\bk{\hat\O\p{t}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the TST expansion and benchmarking methods
when squeezing the initial state $\ket{-\Z}$ rather than $\ket{\X}$:
for initial states $\ket{-\Z}$, all initial-time correlators
$\bk{\hat\S_{\v m}}_{t=0}$ vanish unless $m_+=m_-=0$ (see Appendix
\ref{sec:initial_conditions}), so there is a substantially smaller
number of neglected terms with non-zero contribution to the correlator
$\bk{\hat\O\p{t}}$.

Although the TST expansion breaks down at short times, it has two key
advantages over other methods to compute collective spin correlators.
First, computing spin correlators with the TST expansion is generally
much faster and requires much fewer computing resources than the
alternatives.  The quantum trajectory Monte Carlo simulations
performed for Figure \ref{fig:squeezing_incoherent}, for example, take
$\sim10^4$ CPU hours to compute on standard modern computing hardware;
the bulk of this time is spent performing sparse matrix
multiplication, leaving little room to further optimize run time.
Parallelization can reduce the actual run time of these quantum
trajectory simulations to $\sim10$ hours, but at the cost of greatly
increasing computing resource requirements.  The TST expansion results
in Figure \ref{fig:squeezing_incoherent}, meanwhile, take $\sim10$
seconds to compute with a single CPU on similar hardware.

@@

\begin{figure*}
  \centering
  \subfloat[Two-time correlator
  $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
  Eq.~\eqref{eq:two_time}. \label{fig:two_time}]
  {\includegraphics{two_time.pdf}} \\
  \subfloat[Squared commutator $D\p{t}$ defined in
  Eq.~\eqref{eq:four_point}. \label{fig:four_point}]
  {\includegraphics{OTOC.pdf}}
  \caption{The two-time connected correlator $C\p{t}$ and squared
    commutator $D\p{t}$, respectively defined in
    Eqs.~\eqref{eq:two_time} and \eqref{eq:four_point}, for $N=10^4$
    spins initially in the polarized state
    $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$ evolving under
    the squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.
    Results are shown for time evolution without decoherence (leftmost
    panels), as well as with decoherence at rates $\Gamma_\a$ or
    $\gamma_\a$, respectively corresponding to collective or
    single-spin decoherence through jump operators $\hat S_\a$ or
    $\set{\hat s_\a^{(j)}}$.  Collective decoherence rates
    ($\Gamma_\a$) were chosen to be comparable to the OAT squeezing
    strength $\chi$, while single-spin decoherence rates ($\gamma_\a$)
    were scaled by an additional factor of $S=N/2$ in order to yield
    comparable effects on the correlators $C\p{t}$ and $D\p{t}$ at the
    times shown.  At short times, the multi-time correlations captured
    by $C\p{t}$ and $D\p{t}$ are essentially insensitive to collective
    dephasing ($\Gamma_\z$), while their magnitudes are generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:multi_time}
\end{figure*}

% TODO: in contrast to typical OTOCs, here the operators "almost commute" -- up to 1/N terms

% TODO: mention short-time expansion, see only polynomial growth

% TODO: figures 3a, 3b: 3 panels each; no decoherence, decoherence in fig. 2, and strong decoherence

% TODO: add vertical axis labels for G_- and g_- (?)

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of two-time correlation functions and
out-of-time-ordered correlators (OTOCs).  In particular, we consider
the effect of decoherence on short-time behavior of the two-time
connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\p{\bk{\hat S_+\p{t} \hat S_-\p{0}}
    - \bk{\hat S_+\p{t}}\bk{\hat S_-\p{0}}},
  \label{eq:two_time}
\end{align}
and the expectation value of a squared commutator,
\begin{align}
  D\p{t}
  \equiv \f1{S^2}\Braket{\sp{\hat S_+\p{t},\hat S_-\p{0}}_-^\dag
    \sp{\hat S_+\p{t},\hat S_-\p{0}}_-},
  \label{eq:four_point}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters that diagnose time-crystalline
phases of matter\cite{tucker2018shattered}.  Squared commutators such
as that in Eq.~\eqref{eq:four_point}, meanwhile, are commonly examined
for signatures of quantum chaos and information
scrambling\cite{maldacena2016bound, swingle2018unscrambling,
  garcia-mata2018chaos}.

Figure \ref{fig:multi_time} shows the behavior of $C\p{t}$ and
$D\p{t}$ for $N=10^4$ spins, initially in the state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$, evolving under the
squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both
with and without various sources of decoherence.  At short times, both
correlators are insensitive to collective dephasing, while their
magnitudes are suppressed by other types of decoherence.  The
correlations captured by $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$
also exhibit signatures of the type of decoherence present, with
e.g.~clear dependence of the phase $\phi\p{t}$ on spin decay or
excitation.  Figure \ref{fig:multi_time} serves as an example for the
type of behavior that is accessible at short times with the TST
expansion.  These examples are straightforward extend to equilibrium
settings and spin-boson systems.

@@

Here, we provide a pedagogical tutorial for computing correlators
using the truncated short-time TST expansion.  For concreteness, we
nominally consider $N$ spins evolving under the one-axis twisting
(OAT) Hamiltonian
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
\end{align}
additionally subject to spontaneous single-spin decay at rate
$\gamma_-$, with jump operators $\A_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$.
The equation of motion for a Heisenberg operator
$S_+^\ell S_\z^m S_-^n$ is
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = i\chi\sp{S_\z^2, S_+^\ell S_\z^m S_-^n}_-
  + \gamma_- \D\p{\A_-} \p{S_+^\ell S_\z^m S_-^n},
  \label{eq:tutorial_EOM}
\end{align}
where we can use the results in appendices \ref{sec:general_product}
and \ref{sec:decoherence_single} respectively to evaluate the
commutator $\sp{S_\z^2, S_+^\ell S_\z^m S_-^n}_-$ and dissipator
$\D\p{\A_-} \p{S_+^\ell S_\z^m S_-^n}$ in \eqref{eq:tutorial_EOM} (see
\eqref{eq:general_product} and \eqref{eq:decay_diff} in particular),
giving us
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n} =
  i\chi \p{\ell-n} S_+^\ell \p{\ell+n+2S_\z} S_\z^m S_-^n
  + \gamma_- S_+^\ell \sp{\p{S+S_\z}\p{-1+S_\z}^m
    - \p{S+\f{\ell+n}{2}+S_\z} S_\z^m} S_-^n.
  \label{eq:tutorial_EOM_expanded}
\end{align}
In practice, we do not want to keep track of such an expansion by
hand, especially in the case of e.g.~the two-axis twisting (TAT) and
twist-and-turn (TNT) models with more general types of decoherence,
for which the analogue of \eqref{eq:tutorial_EOM_expanded} may take
several lines just to write out in full.  Defining the operators
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ with
$\v m\equiv\p{m_+,m_\z,m_-}$ for shorthand, we note that the vector
space spanned by $\set{\S_{\v m}}$ is closed under time evolution.  We
therefore expand
\begin{align}
  \f{d}{dt} \S_{\v n} \equiv T \S_{\v n}
  = \sum_{\v m} \S_{\v m} T_{\v m\v n},
  \label{eq:tutorial_EOM_general}
\end{align}
where $T$ is a super-operator that generates time evolution for
Heisenberg operators.  In the present example, the matrix elements
$T_{\v m\v n}\in\C$ of $T$ are defined by
\eqref{eq:tutorial_EOM_expanded} and \eqref{eq:tutorial_EOM_general}.
For any Hamiltonian $H$ with decoherence characterized by sets of jump
operators $\A$ and decoherence rates $\gamma_A$, the matrix elements
$T_{\v m\v n}$ are more generally defined by
\begin{align}
  T \S_{\v n}
  = i\sp{H,\S_{\v n}}_- + \sum_\A \gamma_\A \D\p{\A} \S_{\v n}
  = \sum_{\v m} \S_{\v m} T_{\v m\v n}.
\end{align}
The results in Appendices \ref{sec:general_product},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective} can
be used to write model-agnostic codes that compute matrix elements
$T_{\v m\v n}$, taking a particular Hamiltonian $H$ and decoherence
processes $\set{\p{\A,\gamma_\A}}$ as inputs.

>>>>

\title{Short-time expansion of Heisenberg operators in open collective
  quantum spin systems}

@@

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $\hat S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is
the directionally-unbiased Ramsey squeezing parameter determined by
the maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$
over that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\lvert\braket{\uv S}\rvert^2}
  \min_{\substack{\v v\perp\bk{\uv S}\\\v v\cdot\v v=1}}
    \Braket{\p{\uv S\cdot\v v}^2},
  \label{eq:squeezing}
\end{align}
where $\uv S\equiv\p{\hat S_\x,\hat S_\y,\hat S_\z}$ is a collective
spin operator-valued vector, and the minimization is performed over
all unit vectors $\v v$ orthogonal to the mean spin vector
$\bk{\uv S}$.  This squeezing parameter is entirely determined by one-
and two-spin correlators of the form $\bk{\hat S_\a}$ and
$\bk{\hat S_\a \hat S_\b}$.  For the unitary dynamics discussed in
this work, these correlators are obtainable via exact simulations of
quantum dynamics in the $\p{N+1}$-dimensional Dicke manifold of states
$\set{\ket{S,m}}$ with net spin $S$ and spin projection $m$ onto the
$z$ axis, i.e.~with $\bk{S,m|\uv S^2|S,m}=S\p{S+1}$ and
$\bk{S,m|\hat S_\z|S,m}=m$ for $m\in\set{-S,-S+1,\cdots,S}$.  In the
presence of single-spin or collective decoherence, meanwhile, these
correlators are obtainable with the collective-spin quantum trajectory
Monte Carlo method developed in ref.~\cite{zhang2018montecarlo}.  In
this work, these exact and quantum trajectory simulations will be used
to benchmark the TST expansion in Eq.~\eqref{eq:TST}.

\begin{figure}
  \centering
  \subfloat[Squeezing with unitary dynamics
  \label{fig:squeezing_unitary}]
  {\includegraphics{unitary.pdf}} \\
  \subfloat[Squeezing with decoherence:
  $\gamma_-=\gamma_+=\gamma_\z=\chi$
  \label{fig:squeezing_incoherent}]
  {\includegraphics{decoherence_weak.pdf}}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under \sref{fig:squeezing_unitary} unitary and
    \sref{fig:squeezing_incoherent} non-unitary dynamics, computed
    using exact methods (solid lines), quantum trajectory simulations
    (dots), and the TST expansion in Eq.~\eqref{eq:TST} with $M=35$
    (darker dashed lines).  Solid circles mark the times at which the
    TST expansion gives an unphysical result with $\xi^2<0$.}
  \label{fig:benchmarking}
\end{figure}

Figure \ref{fig:benchmarking} compares the squeezing parameter $\xi^2$
for $N=10^4$ spins initially in the state $\ket{\X}$ evolved under the
Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}, as computed via
both benchmarking simulations and the TST expansion in
Eq.~\eqref{eq:TST} with $M=35$.  Squeezing is shown for both unitary
dynamics (Figure \ref{fig:squeezing_unitary}), as well as non-unitary
dynamics in the presence of spontaneous decay, excitation, and
dephasing of individual spins at rates $\chi$ (Figure
\ref{fig:squeezing_incoherent}), respectively described by the sets of
jump operators $\A_\a\equiv\set{\hat s_\a^{(j)}}$ with corresponding
decoherence rates $\gamma_\a=\chi$ for $\a\in\set{-,+,\z}$.  The
results shown in Figure \ref{fig:benchmarking} were computed in a
rotated basis with
$\p{\hat s_\z,\hat s_\x}\to\p{\hat s_\x,-\hat s_\z}$ and
$\ket\X\to\ket{-\Z}\equiv\ket\dn^{\otimes N}$ (together with
appropriate transformations of the Hamiltonian and jump operators);
the only effect of this rotation on the results presented in Figure
\ref{fig:benchmarking} is to prolong the time for which the TST
expansion agrees with the benchmarking simulations.  The reason for
different results in rotated basis has to do with the breakdown of the
TST expansion, which we discuss further below.

The main lesson from Figure \ref{fig:benchmarking} is that the TST
expansion gives essentially exact results right up until a sudden and
drastic departure that can be diagnosed by inspection.
% todo: comment on diagnosing breakdown by examining individual correlators, rather than squeezing
This drastic departure occurs because the TST expansion neglects
high-weight (i.e.~large-$\abs{\v m}$) operators $\hat\S_{\v m}$ whose
contributions to a Heisenberg operator $\hat\O\p{t}$ of interest
eventually become non-negligible.  For sufficiently large times $t$,
these individual contributions may generally be large compared to the
value of $\bk{\hat\O\p{t}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the TST expansion and benchmarking methods
when squeezing the initial state $\ket{-\Z}$ rather than $\ket{\X}$:
for initial states $\ket{-\Z}$, all initial-time correlators
$\bk{\hat\S_{\v m}}_{t=0}$ vanish unless $m_+=m_-=0$ (see Appendix
\ref{sec:initial_conditions}), so there is a substantially smaller
number of neglected terms with non-zero contribution to the correlator
$\bk{\hat\O\p{t}}$.

Although the TST expansion breaks down at short times, it has two key
advantages over other methods to compute collective spin correlators.
First, computing spin correlators with the TST expansion is generally
much faster and requires much fewer computing resources than the
alternatives.  The quantum trajectory Monte Carlo simulations
performed for Figure \ref{fig:squeezing_incoherent}, for example, take
$\sim10^4$ CPU hours to compute on standard modern computing hardware;
the bulk of this time is spent performing sparse matrix-vector
multiplication, leaving little room to further optimize run time.
Parallelization can reduce the actual run time of these quantum
trajectory simulations to $\sim10$ hours, but at the cost of greatly
increasing computing resource requirements.  The TST expansion results
in Figure \ref{fig:squeezing_incoherent}, meanwhile, take $\sim10$
seconds to compute with a single CPU on similar hardware.

@@

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of two-time correlation functions and
out-of-time-ordered correlators (OTOCs).  In particular, we consider
the effect of decoherence on short-time behavior of the two-time
connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\p{\bk{\hat S_+\p{t} \hat S_-\p{0}}
    - \bk{\hat S_+\p{t}}\bk{\hat S_-\p{0}}},
  \label{eq:two_time}
\end{align}
and the expectation value of a squared commutator,
\begin{align}
  D\p{t}
  \equiv \f1{S^2}\Braket{\sp{\hat S_+\p{t},\hat S_-\p{0}}_-^\dag
    \sp{\hat S_+\p{t},\hat S_-\p{0}}_-},
  \label{eq:four_point}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters that diagnose time-crystalline
phases of matter\cite{tucker2018shattered}.  Squared commutators such
as that in Eq.~\eqref{eq:four_point}, meanwhile, are commonly examined
for signatures of quantum chaos and information
scrambling\cite{maldacena2016bound, swingle2018unscrambling,
  garcia-mata2018chaos}.  In typical scenarios, such squared
commutators initially vanish by construction due to a choice of
operators with spatially separated non-trivial support.  Collective
spin systems, however, have no intrinsic notion of locality or spatial
separation.  In our case, therefore, with the choice of initial state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$ we merely have
$D\p{0}\sim1/N$.

\begin{figure}
  \centering
  \subfloat[Two-time correlator
  $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$. \label{fig:two_time}]
  {\includegraphics{two_time.pdf}} \\
  \subfloat[Squared commutator $D\p{t}$. \label{fig:four_point}]
  {\includegraphics{FPC.pdf}}
  \caption{The two-time connected correlator $C\p{t}$ and squared
    commutator $D\p{t}$, respectively defined in
    Eqs.~\eqref{eq:two_time} and \eqref{eq:four_point}, for $N=10^4$
    spins initially in the polarized state
    $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$ evolving under
    the squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.
    Results are shown for both unitary dynamics (solid lines) and
    non-unitary dynamics with $\gamma_-=\gamma_+=\gamma_\z=100\chi$
    (dashed lines), computed using the TST expansion in
    Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:multi_time}
\end{figure}

% todo: add commentary/discussion about growth of C(t)

Figure \ref{fig:multi_time} shows the behavior of $C\p{t}$ and
$D\p{t}$ for $N=10^4$ spins, initially in the state $\ket\X$, evolving
under the squeezing Hamiltonians in
Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both with and without spontaneous
decay, excitation, and dephasing of individual spins at rates
$\gamma_-=\gamma_+=\gamma_\z=100\chi$.  In the case of TAT and TNT,
after an initial ramp-up period these correlators grow exponentially
as $\sim\exp\p{a\chi t}$ with $a\sim N$.  The growth of these
correlators is definitively sub-exponential for OAT.  In the case of
unitary time evolution under OAT in particular, to an excellent
approximation we find that $D\p{t}=D\p{0}+\p{b\chi t}^2$ with
$b=N+O\p{1}$.  As may be expected, the growth rates of $C\p{t}$ and
$D\p{t}$ are generally suppressed by decoherence.  Figure
\ref{fig:multi_time} serves as an example for the type of behavior
that is accessible at short times with the TST expansion.  These
examples are straightforward extend to equilibrium settings and
spin-boson systems.

@@

Here, we provide a pedagogical tutorial for computing correlators
using the truncated short-time TST expansion.  For concreteness, we
nominally consider $N$ spins evolving under the one-axis twisting
(OAT) Hamiltonian
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
\end{align}
additionally subject to spontaneous single-spin decay at rate
$\gamma_-$, with jump operators $\A_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$.
The equation of motion for a Heisenberg operator
$S_+^\ell S_\z^m S_-^n$ is
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = i\chi\sp{S_\z^2, S_+^\ell S_\z^m S_-^n}_-
  + \gamma_- \D\p{\A_-} \p{S_+^\ell S_\z^m S_-^n},
  \label{eq:tutorial_EOM}
\end{align}
where we can use the results in appendices \ref{sec:general_product}
and \ref{sec:decoherence_single} respectively to evaluate the
commutator $\sp{S_\z^2, S_+^\ell S_\z^m S_-^n}_-$ and dissipator
$\D\p{\A_-} \p{S_+^\ell S_\z^m S_-^n}$ in \eqref{eq:tutorial_EOM} (see
\eqref{eq:general_product} and \eqref{eq:decay_diff} in particular),
giving us
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n} =
  i\chi \p{\ell-n} S_+^\ell \p{\ell+n+2S_\z} S_\z^m S_-^n
  + \gamma_- S_+^\ell \sp{\p{S+S_\z}\p{-1+S_\z}^m
    - \p{S+\f{\ell+n}{2}+S_\z} S_\z^m} S_-^n.
  \label{eq:tutorial_EOM_expanded}
\end{align}
In practice, we do not want to keep track of such an expansion by
hand, especially in the case of e.g.~the two-axis twisting (TAT) and
twist-and-turn (TNT) models with more general types of decoherence,
for which the analogue of \eqref{eq:tutorial_EOM_expanded} may take
several lines just to write out in full.  Defining the operators
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ with
$\v m\equiv\p{m_+,m_\z,m_-}$ for shorthand, we note that the vector
space spanned by $\set{\S_{\v m}}$ is closed under time evolution.  We
therefore expand
\begin{align}
  \f{d}{dt} \S_{\v n} \equiv T \S_{\v n}
  = \sum_{\v m} \S_{\v m} T_{\v m\v n},
  \label{eq:tutorial_EOM_general}
\end{align}
where $T$ is a super-operator that generates time evolution for
Heisenberg operators.  In the present example, the matrix elements
$T_{\v m\v n}\in\C$ of $T$ are defined by
\eqref{eq:tutorial_EOM_expanded} and \eqref{eq:tutorial_EOM_general}.
For any Hamiltonian $H$ with decoherence characterized by sets of jump
operators $\A$ and decoherence rates $\gamma_\A$, the matrix elements
$T_{\v m\v n}$ are more generally defined by
\begin{align}
  T \S_{\v n}
  = i\sp{H,\S_{\v n}}_- + \sum_\A \gamma_\A \D\p{\A} \S_{\v n}
  = \sum_{\v m} \S_{\v m} T_{\v m\v n}.
\end{align}
The results in Appendices \ref{sec:general_product},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective} can
be used to write model-agnostic codes that compute matrix elements
$T_{\v m\v n}$, taking a particular Hamiltonian $H$ and decoherence
processes $\set{\p{\A,\gamma_\A}}$ as inputs.
